### 2014-07-24 Studies/2012_Chicago/Report/Chicago_Fire.tex
Chicago: update report, input files, and figures

\chapter{Fire Incident Summary}
\label{fire_sum}
The following account of events is based on information provided from the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report~\cite{NIOSH:Bowyer}. Details regarding times are taken from the report and should be considered approximate values.

On November 2, 2012, a fire in a 2 1/2 story wood frame residential structure resulted in the death of a fire captain and injury to a fire fighter. At approximately 17:16 hours, a dispatch was made for ``smoke in the area''. The dispatch was made because a person on the street saw smoke and notified the residents, who all promptly left the structure. A battalion chief (BC19) from the nearest station arrived on-scene about one minute after dispatch; the station was only several blocks away. BC19 performed an exterior size-up and reported a working fire with heavy smoke coming from the front and rear of the structure's attic.  BC19 assumed incident command (IC) of the fire scene. After speaking with the building occupants, the IC confirmed the structure was evacuated. The IC entered the structure to investigate the second floor. He found only light smoke on the second floor and a glow around the doorway that led to enclosed rear porch.

Engine 123 (E123) arrived on scene at approximately 17:18. The IC met the E123 captain and firefighter at the front door to the structure. The captain and firefighter took their hose line to the second floor of the structure. At approximately 17:21, the captain and firefighter were on the second floor with their hoseline and had begun advancing the charged line toward the rear second floor door between the apartment and the enclosed porch. Approximately 30 seconds later, a fire fighter from Tower Ladder 39 (TL39) enters the enclosed porch on the rear of the structure by opening the rear exterior door to the first floor of the porch. The firefighter from TL39, notices ``the fire light up'' in the stairwell of the enclosed porch, and then kicks in the first level door between the rear porch and the main structure. The firefighter did not see any fire in the main part of the structure and proceeded to exit the structure. Given the amount of fire showing from the second floor of the rear porch and ``heavy fire'' in the stairwell of the rear porch, the BC radios the E123 Captain at 17:23, to alert him to the fire conditions and to advise him that E49 would be applying water to the fire on the rear of the structure from the exterior. No response was made by the E123 Captain. It is believed that the door between the rear porch and the second floor hallway, where the E123 Captain and firefighter were located may have failed by this time.  Approximately a minute later E49 applied water to the fire in the attic.

Squad 5 (S5) entered the front of the structure at approximately 17:25 and headed to the second floor. They heard a commotion and found the E123 Captain collapsed on the kitchen floor and entangled with the E123 firefighter. A Mayday was called on the radio by S5 at approximately 17:27 and the E123 Captain was removed from the front of the building at approximately 17:29 and succumbed to his injuries at the hospital at 17:38. Table~\ref{tab:fire_info} presents the timeline of the fire.

\begin{table}
\centering
\captionof{table}{Abridged NISOH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{l l}
\toprule[1.5pt]
Incident Time [hr:min:s] & Fire Behavior / Fireground Operation  \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm $-$ two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic self ventilation - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 Capt. and FF on 2$^{nd}$ floor with 1 3/4 in hose line. \\[.25cm]
17:21:30 & 1$^{st}$ floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123: Capt. or FF.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 Capt. carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Timeline}
\label{timeline}
Table~\ref{tab:firesim} shows the fireground timeline along with the corresponding simulation times.

\begin{table}
\centering
\captionof{table}{Fire - Simulation Event Timeline}\label{tab:firesim}
\begin{tabular}{l l l}
\toprule[1.5pt]
Incident Time [hr:min:s] & Model Time [s] & Fire Behavior / Fireground Operation  \\
\midrule
\multirow{2}{*}{17:15:00} &    & \multirow{2}{*}{\parbox{9cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00} & \multirow{2}{*}{}   &  \multirow{2}{*}{\parbox{9cm} {Dispatch for a ``Still'' alarm $-$ two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
         &  \multirow{2}{*} {0} &  \multirow{2}{*}{\parbox{9cm} {Simulation starts. Attic fire ignites. Initial self ventilation on charlie side. Front door open.}} \\
         & & \\[0.25cm]
\multirow{3}{*}{17:20:00} & \multirow{3}{*}{40}  & \multirow{3}{*}{\parbox{9cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear. Rear porch ignites in model.}} \\
         & & \\[.25cm]
         & & \\[.25cm]  
         & 45 & Ventilation through wall penetration on rear of structure. \\[.25cm] 
         & 55 & Ventilation through wall penetration on left side of structure. \\[.25cm] 
17:21:00 & 100 & E123 Capt. and FF on 2$^{nd}$ floor with 1 3/4-in hose line. \\[.25cm]
\multirow{2}{*}{17:21:30} & \multirow{2}{*}{130} &  \multirow{2}{*}{\parbox{9cm}{1$^{st}$ floor exterior porch door opened. Porch fire ignites in FDS model.}} \\
        &  & \\[.25cm]
        & 135 & Additional ventilation through left side of structure. \\[.25cm] 
        & 160 & Top of interior door fails in FDS model. \\[.25cm] 
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{220} & \multirow{2}{*}{\parbox{9cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123: Capt. or FF.}} \\
         & & \\[.25cm]
17:24:00    &   & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    &   & SQ5 makes entry through front of structure. \\[.25cm]
\multirow{2}{*}{17:27:00}    & \multirow{2}{*}{300}  & \multirow{2}{*}{\parbox{9cm} {After hearing a Mayday call from SQ5, IC calls a Mayday. Simulation ends.}}\\
         & &  \\[.25cm]
17:29:00    &   & E123 Capt. carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

@@

\section{Hallway Flow Path}
From on-scene reports from the BC (Section~\ref{fire_sum}) and simulation results (Section~\ref{temp}), interior conditions on the second floor of the structure were initially tenable. From the NIOSH report, while the BC was inside the structure on the second floor, the  BC observed only a light haze and a glow around the door between the hallway and enclosed porch. This indicated that some amount of the fire was burring in the area of the enclosed porch. The simulation shows a high hazard area in the attic and the enclosed porch at this time. By the time the E123 Captain and FF arrived on the second floor (Table~\ref{tab:firesim}) and approached the rear of the structure, the rear porch was fully involved. Temperature and pressure, greater than 260~$^{\circ}$C (500~$^{\circ}$F) and 5~Pa respectively, were on the porch side of the steel door at the rear of the structure. Exposure to these conditions weakened the structural integrity of the door, eventually leading to a partial failure. Fig.~(\ref{fig:chicago_doorfold}) shows the state of the rear door post fire. Note the large crease in the door just above the knob.

@@

While we have estimated the time of failure for the door, the key take away is that after the door failed an interior flow path from the enclosed porch to the kitchen of the structure was created. This resulted in a rapid change in the conditions within the flow path. The combustion-gases at elevated temperature and pressure  in the enclosed porch were vented by the failure of the door to lower pressure areas in the hallways and kitchen, where the conditions seems to equalize. The arrows in Fig.~(\ref{fig:flowpath_1}) show the flow direction of gases on the second floor following the door failure.

@@

A person is susceptible to second degree burn injuries when exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F) \cite{designation2003c}. While firefighters wear protective gear, gear only offers finite levels of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F) \cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) \cite{nfpa2013standard}. Extended exposure to elevated temperatures can result in significant heat transferred to the firefighter, putting him/her at risk. Therefore, for analysis of the modeled exposure temperatures, we set the maximum visualized temperature to be all values that are 260~$^{\circ}$C (500~$^{\circ}$F) or above. This means that temperatures visualized in the highest range in the model output are at a minimum of 260~$^{\circ}$C (500~$^{\circ}$F); predicted temperatures could in fact be higher.

Prior to door failure, the elevated temperatures on the second floor were limited to the rear porch, Fig.~(\ref{fig:temp_top_159}). Temperatures are highest at the back-left corner where there is sufficient oxygen due to the section of porch wall that had burned away allowing additional oxygen to mix with the fuel rich fire gases. Five seconds after the door fails, hot gases have traveled the length of the interior hallway and approached the doorway into the kitchen, approximately 4.5~m (14.8~ft) away from the source fire (cf. Fig.~(\ref{fig:temp_top_165})). Within one minute after the door fails, the hallway temperatures are nominally at or above 260~$^{\circ}$C (500~$^{\circ}$F) through out the hallway (cf Fig.~(\ref{fig:temp_top_220})). This illustrates that the high temperature gases seen five seconds after failure were not just a burst but rather the front of a sustained exhaust gas flow from the fire burning in the enclosed rear porch. The conditions in this hallway changed from tenable to high-hazard very rapidly following the door failure. Fig.~(\ref{fig:temp_top_220}) also shows that the temperatures in the kitchen were at or below 90~$^{\circ}$C (194~$^{\circ}$F) and temperatures at the top of the stairs were at or below 70~$^{\circ}$C (158~$^{\circ}$F). At these points in the structure, the exhaust gases in the flow path are air cooled/diluted by the inflow ambient gas. These modeled temperatures are consistent with the post-incident conditions documented in the kitchen and second floor stairwell.
\begin{figure}[h!]
\centering
\includegraphics[width=.7\textwidth]{../Figures/west_50th_baseline_top_159_6ft}
\input{../Scripts/colorbar_std} 
\input{../Scripts/colorbar_metric}
\caption{Temperature contours on the second floor, 1 second prior to door failure, 1.83~m (6~ft) above the floor.}
\label{fig:temp_top_159}
\end{figure}

@@

\chapter{Summary}
NIST's Fire Dynamics Simulator was used to provide insight into the fire dynamics within a 2 1/2 story single family structure in Chicago, Il that resulted in the death of a firefighter. The modeled fuel, fire size, and fire growth rate were estimated taking into account all available information: NIOSH post-incident report~\cite{NIOSH:Bowyer}, post-incident pictures, videos, and relevant literature. This resulted in a specified source fire of 12.3~MW, however based on the limited ventilation in the attic and enclosed porch the model generated a peak HRR of approximately 10 MW as a result of the limited supply of oxygen.  The fire began in the attic and spread down into the enclosed porch on the rear (Charlie side) of the structure.  The spread of heat and fire gases into the second floor of the structure was limited by a closed steel faced, wood framed door that was located between the enclosed porch and the hallway.  After exposure to the elevated temperature and pressure from the fire in the porch area, the wood frame of the door decomposed and the steel faces of the door failed and collapsed inward.  The door failure resulted in the creation of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor.  The temperature of the gases that flowed into the hallway at face height exceeded 260~$^{\circ}$C (500~$^{\circ}$F).

Unfortunately a FF and a Captain were advancing a hose line in the hallway at the time of the failure of the door.  The Captain was overcome by the rapid change in conditions before he could exit to the safety of the kitchen area.  After a Mayday, the Captain was rescued from the structure but later succumbed to his injuries.

>>>>

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident follows the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For analysis purposes, details regarding times are considered approximate values. The following incident narrative is taken from that NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' \footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} alarm for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure?s attic. Per fire department standard operating procedures, dispatch initiated a RIT \footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.} response. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2nd floor. He noticed only a light haze throughout and glow around the Sector 3 door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a horseshoe load which is 100 feet of 2-$\tfrac{1}{2}$ inch, a gated wye, and 150 feet of 1-$\tfrac{3}{4}$ inch hoseline to go to the 2nd floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1st floor apartment and worked his way to [the rear of the structure]. He opened the back door to the covered porch and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2nd floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2-$\tfrac{1}{2}$ inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [on the left side]. A TL39 firefighter went to [rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door to the first floor apartment, preceded to walk in, saw no fire then backed out.

At 1723 hours, the IC radioed the [E123 captain] that there was heavy fire in the covered porch and attic area and that E49 was going to put water on the fire, around the [rear] attic window, but there was no acknowledgement from the [captain]. E49 proceeded to put water on the [rear of the structure]. The IC returned out front to the command post and donned his turnout gear. The lieutenant from TL39, the lieutenant and 2 firefighters from SQ5, and a E123 firefighter/paramedic (FF/PM) were near the kitchen area on the 2nd floor when they heard a loud commotion. The FF/PM heard the [E123 captain] yell ``get out of here.'' The FF/PM had no radio and he couldn't locate the [captain's] radio so he yelled Mayday as he tried to get the victim and other crew member untangled. The TL39 Lieutenant and SQ5 fire fighters heard the FF/PM's verbal Mayday and the SQ5 Lieutenant tried to transmit a Mayday over heavy radio traffic.

At 1727 hours, the TL39 crew had just completed the first hole in the roof about midway on [the right side] roof with minimal fire showing, when they heard the Mayday. The IC verified that a firefighter was down, called a Mayday, and requested a 2-11 Assignment \footnote{A 2-11 alarm is a 2nd alarm upgrade. The response is four engines, two trucks, one tower ladder, two battalion chiefs, one deputy district chief, one air mask service bottle truck, and a media affairs unit.}. Dispatch initiated a 2-11 alarm. SQ5 and other members on the 2nd floor grabbed the [E123 captain] and got him down the stairs. The TL39 crew, with assistance from a T33 and SQ5 firefighter, had just completed a second hole on [right side] of the roof about a third of the way back when conditions worsened. At 1729 hours, the roof ventilation crew was back in the aerial basket when they noticed the victim being brought out to the front yard. The victim was nonresponsive in the front yard and CPR was successfully performed. The IC met the A19 crew and escorted them to the [E123 captain]. The revived [captain] was responsive and talking to the paramedics as he was loaded into the ambulance. At 1738 hours, the [captain] was transported to the local hospital where he had complications during airway management and died.
\end{quote}

\noindent For an overview of timeline of events, see Table~\ref{tab:fire_info}. A plan view of the structure and location where the captain collapsed can be found in Fig.~(\ref{fig:simp_geom}).
\begin{table}
\centering
\captionof{table}{Abridged NISOH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{l l}
\toprule[1.5pt]
Incident Time [hr:min:s] & Fire Behavior / Fireground Operation  \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm $-$ two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on 2nd floor with 1-$\tfrac{3}{4}$ inch hoseline. \\[.25cm]
17:21:30 & 1st floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123: captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

@@

To examine results from the model, it is important to link the timelines from fire scene to the simulation. Table~\ref{tab:firesim} shows the fireground timeline along with the corresponding simulation times.

\begin{table}
\centering
\captionof{table}{Fire - Simulation Event Timeline}\label{tab:firesim}
\begin{tabular}{l l l}
\toprule[1.5pt]
Incident Time [hr:min:s] & Model Time [s] & Fire Behavior / Fireground Operation  \\
\midrule
\multirow{2}{*}{17:15:00} &    & \multirow{2}{*}{\parbox{9cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00} & \multirow{2}{*}{}   &  \multirow{2}{*}{\parbox{9cm} {Dispatch for a ``Still'' alarm $-$ two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
         &  \multirow{2}{*} {0} &  \multirow{2}{*}{\parbox{9cm} {Simulation starts. Attic fire ignites. Initial ventilation through rear of structure. Front door open.}} \\
         & & \\[0.25cm]
\multirow{3}{*}{17:20:00} & \multirow{3}{*}{40}  & \multirow{3}{*}{\parbox{9cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear. Rear porch ignites in model.}} \\
         & & \\[.25cm]
         & & \\[.25cm]  
         & 45 & Ventilation through wall penetration on rear of structure. \\[.25cm] 
         & 55 & Ventilation through wall penetration on left side of structure. \\[.25cm] 
17:21:00 & 100 & E123 captain and firefighter on 2nd floor with 1-$\tfrac{3}{4}$ inch hoseline. \\[.25cm]
\multirow{2}{*}{17:21:30} & \multirow{2}{*}{130} &  \multirow{2}{*}{\parbox{9cm}{1st floor exterior porch door opened. Porch fire ignites in FDS model.}} \\
        &  & \\[.25cm]
        & 135 & Additional ventilation through left side of structure. \\[.25cm] 
        & 160 & Top of interior door fails in FDS model. \\[.25cm] 
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{220} & \multirow{2}{*}{\parbox{9cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123: captain or firefighter.}} \\
         & & \\[.25cm]
17:24:00    &   & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    &   & SQ5 makes entry through front of structure. \\[.25cm]
\multirow{2}{*}{17:27:00}    & \multirow{2}{*}{300}  & \multirow{2}{*}{\parbox{9cm} {After hearing a Mayday call from SQ5, IC calls a Mayday. Simulation ends.}}\\
         & &  \\[.25cm]
17:29:00    &   & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Hallway Flow Path}
From on-scene reports (Section~\ref{fire_sum}) and simulation results (Section~\ref{temp}), interior conditions on the second floor of the structure were initially tenable. From the NIOSH report, while the IC was inside the structure on the second floor, only a light haze and a glow around the door between the hallway and enclosed porch was observed. This indicated fthe presence of fire in the area of the enclosed porch. The simulation shows a high hazard area in the attic and the enclosed porch at this time. By the time the E123 captain and firefighter arrived on the second floor (Table~\ref{tab:firesim}) and approached the rear of the structure, the rear porch was fully involved. Temperature and pressure, greater than 260~$^{\circ}$C (500~$^{\circ}$F) and 5~Pa respectively, were on the porch side of the steel door at the rear of the structure. Exposure to these conditions weakened the structural integrity of the door, which may have lead to a partial failure. Fig.~(\ref{fig:chicago_doorfold}) shows the state of the rear door post fire. Note the large crease in the door just above the knob.

@@

While we had to estimate the time of failure for the door, the key take away is that after the door failed an interior flow path from the enclosed porch to the kitchen of the structure was created. This resulted in a rapid change in the conditions within the flow path. The combustion-gases at elevated temperature and pressure in the enclosed porch were vented by the failure of the door to lower pressure areas in the hallways and kitchen, where the conditions seemed to equalize. The arrows in Fig.~(\ref{fig:flowpath_1}) show the flow direction of gases on the second floor following the door failure.

@@

A person is susceptible to second degree burn injuries when exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F) \cite{designation2003c}. While firefighters wear protective gear, gear only offers finite levels of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F) \cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) \cite{nfpa2013standard}. Extended exposure to elevated temperatures can result in significant heat transferred to the firefighter, putting him/her at risk. Therefore, for analysis of the modeled exposure temperatures, we set the maximum visualized temperature to be all values that are 260~$^{\circ}$C (500~$^{\circ}$F) or above. This means that temperatures visualized in the highest range in the model output are at a minimum of 260~$^{\circ}$C (500~$^{\circ}$F); simulated temperatures could in fact be higher.

Prior to the door failure, the elevated temperatures on the second floor were limited to the rear porch, Fig.~(\ref{fig:temp_top_159}). Temperatures are highest at the back-left corner where there is sufficient oxygen due to the section of porch wall that had burned away allowing additional oxygen to mix with the fuel rich fire gases. Five seconds after the door fails, hot gases have traveled the length of the interior hallway and approached the doorway into the kitchen, approximately 4.5~m (14.8~ft) away from the source fire (cf. Fig.~(\ref{fig:temp_top_165})). Within one minute after the door fails, the hallway temperatures are nominally at or above 260~$^{\circ}$C (500~$^{\circ}$F) throughout the hallway (cf Fig.~(\ref{fig:temp_top_220})). This illustrates that the high temperature gases seen five seconds after failure were not just a burst but rather the front of a sustained exhaust gas flow from the fire burning in the enclosed rear porch. The conditions in this hallway changed from tenable to high-hazard very rapidly following the door failure. Fig.~(\ref{fig:temp_top_220}) also shows that the temperatures in the kitchen were at or below 90~$^{\circ}$C (194~$^{\circ}$F) and temperatures at the top of the stairs were at or below 70~$^{\circ}$C (158~$^{\circ}$F). At these points in the structure, the exhaust gases in the flow path are air cooled/diluted by the inflow ambient gas. These modeled temperatures are consistent with the post-incident conditions documented in the kitchen and second floor stairwell.
\begin{figure}[h!]
\centering
\includegraphics[width=.7\textwidth]{../Figures/west_50th_baseline_top_159_6ft}
\input{../Scripts/colorbar_std} 
\input{../Scripts/colorbar_metric}
\caption{Temperature contours on the second floor, 1 second prior to door failure, 1.83~m (6~ft) above the floor.}
\label{fig:temp_top_159}
\end{figure}

@@

\chapter{Summary}
NIST's Fire Dynamics Simulator was used to provide insight into the fire dynamics within a 2 1/2 story single family structure in Chicago, Il that resulted in the death of a firefighter. The modeled fuel, fire size, and fire growth rate were estimated taking into account all available information: NIOSH post-incident report~\cite{NIOSH:Bowyer}, post-incident pictures, videos, and relevant literature. This resulted in a specified source fire of 12.3~MW, however based on the limited ventilation in the attic and enclosed porch the model generated a peak HRR of approximately 10 MW as a result of the limited supply of oxygen.  The fire began in the attic and spread down into the enclosed porch on the rear of the structure.  The spread of heat and fire gases into the second floor of the structure was limited by a closed steel faced, wood framed door that was located between the enclosed porch and the hallway.  After exposure to the elevated temperature and pressure from the fire in the porch area, the wood frame of the door decomposed and the steel faces of the door failed and collapsed inward.  The door failure resulted in the creation of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor.  The temperature of the gases that flowed into the hallway at face height exceeded 260~$^{\circ}$C (500~$^{\circ}$F).

Unfortunately a captain and firefighter were advancing a hose line in the hallway at the time of the failure of the door.  The captain was overcome by the rapid change in conditions before he could exit to the safety of the kitchen area.  After a Mayday, the captain was rescued from the structure but later succumbed to his injuries.
